---
output: github_document
always_allow_html: true
editor_options:
  markdown:
    wrap: 72
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  message = FALSE,
  warning = FALSE,
  fig.retina = 2,
  fig.align = 'center'
)
```

# jmpinput

<!-- badges: start -->

[![License: CC BY
4.0](https://img.shields.io/badge/License-CC_BY_4.0-lightgrey.svg)](https://creativecommons.org/licenses/by/4.0/)

<!-- badges: end -->

The goal of jmpinput is to make the data behind the WHO/UNICEF Joint
Monitoring Programme for Water Supply, Sanitation and Hygiene (JMP)
available as tidy, documented R datasets. The package contains two
data resources from the JMP 2025 release, both prepared by the update
pipeline documented in `data-raw/UPDATING.md`:

- `jmpraw`: the survey-level input data on sanitation that underlies
  the JMP estimates, extracted from the country files published on
  [washdata.org](https://washdata.org).
- `jmpindicators`: the modelled JMP indicator estimates for drinking
  water, sanitation and hygiene from the JMP world file, with variable
  codes normalized to the legacy JMP vocabulary so snapshots stay
  comparable across releases.

## Installation

You can install the development version of jmpinput from
[GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("openwashdata/jmpinput")
```

```{r}
## Run the following code in console if you don't have the packages
## install.packages(c("dplyr", "knitr", "readr", "stringr", "gt", "kableExtra"))
library(dplyr)
library(knitr)
library(readr)
library(stringr)
library(gt)
library(kableExtra)
```

Alternatively, you can download the individual datasets from the table
below. `jmpraw` is available as CSV and XLSX; `jmpindicators` is
available as a gzip-compressed CSV only, because the uncompressed file
is about 38 MB (R's `readr::read_csv()` and most spreadsheet tools
open `.csv.gz` files directly).

1.  Click a download link. A window opens that displays the file in
    your browser, or downloads it directly.
2.  If the file is displayed, right-click anywhere inside the window
    and select "Save Page As...".
3.  Save the file in a folder of your choice.

```{r, echo=FALSE, message=FALSE, warning=FALSE}

extdata_path <- "https://github.com/openwashdata/jmpinput/raw/main/inst/extdata/"

tibble::tribble(
  ~dataset, ~CSV, ~XLSX,
  "jmpraw",
  paste0("[Download CSV](", extdata_path, "jmpraw.csv)"),
  paste0("[Download XLSX](", extdata_path, "jmpraw.xlsx)"),
  "jmpindicators",
  paste0("[Download CSV (gzip)](", extdata_path, "jmpindicators.csv.gz)"),
  ""
) |>
  knitr::kable()

```

## Data

The package provides access to the input data and the modelled
indicator estimates of the WHO/UNICEF Joint Monitoring Programme for
Water Supply, Sanitation and Hygiene (JMP), packaged from the JMP 2025
release (pulled 2026-08-20).

```{r}
library(jmpinput)
```

### jmpraw

The dataset `jmpraw` contains the survey, census and administrative
data points on sanitation that underlie the JMP estimates, one row per
country, source, collection year and sanitation variable. It has
`r nrow(jmpraw)` observations and `r ncol(jmpraw)` variables.

```{r}
jmpraw |>
  head(3) |>
  gt::gt() |>
  gt::as_raw_html()
```

For an overview of the variable names, see the following table.

```{r echo=FALSE, message=FALSE, warning=FALSE}
readr::read_csv("data-raw/dictionary.csv") |>
  dplyr::filter(file_name == "jmpraw.rda") |>
  dplyr::select(variable_name:description) |>
  knitr::kable() |>
  kableExtra::kable_styling("striped") |>
  kableExtra::scroll_box(height = "200px")
```

### jmpindicators

The dataset `jmpindicators` contains the modelled JMP indicator
estimates for drinking water, sanitation and hygiene, one row per
country, estimate year, residence level and indicator. It has
`r nrow(jmpindicators)` observations and `r ncol(jmpindicators)`
variables.

```{r}
jmpindicators |>
  head(3) |>
  gt::gt() |>
  gt::as_raw_html()
```

For an overview of the variable names, see the following table.

```{r echo=FALSE, message=FALSE, warning=FALSE}
readr::read_csv("data-raw/dictionary.csv") |>
  dplyr::filter(file_name == "jmpindicators.rda") |>
  dplyr::select(variable_name:description) |>
  knitr::kable() |>
  kableExtra::kable_styling("striped") |>
  kableExtra::scroll_box(height = "200px")
```

## Example

How many survey-level sanitation data points does each JMP release
hold for Uganda, by source type?

```{r}
library(jmpinput)
library(dplyr)

jmpraw |>
  filter(iso3 == "UGA") |>
  count(type, sort = TRUE)
```

Safely managed sanitation in Uganda over time, from the modelled
estimates:

```{r}
jmpindicators |>
  filter(iso3 == "UGA",
         indicator == "Safely managed",
         service == "sanitation",
         residence == "national",
         year %in% c(2000, 2010, 2020, 2024)) |>
  select(year, percent)
```

## License

Data are available as
[CC-BY](https://github.com/openwashdata/jmpinput/blob/main/LICENSE.md).

## Citation

Please cite this package using:

```{r}
citation("jmpinput")
```
