---
title: "JMP raw data visualisations"
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  out.width = "100%"
)

library(tidyverse)
library(tmap)
library(sf)

jmpraw <- jmpinput::jmpraw %>%
    filter(year >= 2000, value >= 0)


top10 <- function(data) {
    data  %>%
        arrange(desc(n)) %>%
        head(10)  %>%
        mutate(country = countrycode::countrycode(iso3, origin = "iso3c", destination = "country.name")) %>%
        select(country, n)
}


```

# Explore data

Datasets included in the JMP database include:

- Censuses, which in principle collect basic data from all people living within a country. Censuses are always led by national statistical offices.
- Household surveys, which collect data from a subset of households. These may target national, rural, or urban populations, or more limited project or sub-national areas. An appropriate sample design is necessary for survey results to be representative, and surveys are often led by or reviewed and approved by national statistical organizations.
- Administrative data, may consist of information collected by government or non-government entities involved in the delivery or oversight of services. Examples include: water and sanitation inventories and databases, and reports of regulators.
- Other datasets may be available such as compilations by international or regional initiatives (e.g. IB-NET), studies conducted by research institutes, or technical advice received during country consultations.


```{r}

## 5 NAs are countries without data

jmpraw %>%
    count(type)

jmpraw %>%
    count(type, source) %>%
    arrange(desc(n))

jmpraw %>%
    count(san_service_chain)

jmpraw %>%
    ggplot(aes(x = year, fill = type)) +
    geom_bar()

jmpraw %>%
    ggplot(aes(x = year, fill = san_service_chain)) +
    geom_bar()

```

```{r}

jmpraw %>%
    filter(!is.na(san_service_chain)) %>%
    group_by(san_service_chain) %>%
    count() %>%
    ggplot(aes(x = san_service_chain, y = n)) +
    geom_col() +
    labs(
        title = "Number of data sources along sanitation service chain",
        subtitle = "For 232 countries and since 2000"
    ) +
    theme_minimal()

```


```{r}

jmpraw %>%
    filter(!is.na(san_service_chain)) %>%
    count(san_service_chain) %>%
    mutate(
        prop = n / sum(n) * 100
    ) %>%

    ggplot(aes(x = san_service_chain, y = prop)) +
    geom_col() +
    labs(
        title = "Proportion of data sources along sanitation service chain",
        subtitle = "For 232 countries and since 2000"
    )


```


```{r}

jmpraw %>%
    filter(!is.na(san_service_chain)) %>%
    count(year, san_service_chain) %>%
    group_by(year) %>%
    mutate(
        prop = n / sum(n) * 100
    ) %>%

    ggplot(aes(x = year, y = prop, fill = san_service_chain)) +
    geom_col() +
    labs(
        x = NULL,
        y = NULL,
        title = "Proportion of data points along collected variables",
        subtitle = "For all countries since 2000",
        fill = "Variable"
    ) +
    scale_fill_brewer(palette = "BrBG") +
    theme_minimal(base_size = 18) +
    theme(panel.grid.minor = element_blank())


```


```{r}

jmp_iso3_frequency <- jmpraw %>%
    count(iso3)

jmp_iso3_user_interface_2015 <- jmpraw %>%
    filter(year >= 2015) %>%
    filter(san_service_chain == "user interface") %>%
    count(iso3, san_service_chain)

jmp_iso3_emptying_2015 <- jmpraw %>%
    filter(year >= 2015) %>%
    filter(san_service_chain == "emptying") %>%
    count(iso3, san_service_chain)

jmp_iso3_ww_treatment_2015 <- jmpraw %>%
    filter(year >= 2015) %>%
    filter(san_service_chain == "WW treatment") %>%
    count(iso3, san_service_chain)

jmp_iso3_fs_treatment_2015 <- jmpraw %>%
    filter(year >= 2015) %>%
    filter(san_service_chain == "FS treatment") %>%
    count(iso3, san_service_chain)


```


# Maps

```{r}
## prepare world maps
## https://r-tmap.github.io/tmap/articles/basics_modes

data("World")
world_moll = st_transform(World, crs = "+proj=moll")

tmap_mode("plot")

```


## All data points since 2000

```{r}
world_moll %>%
    left_join(jmp_iso3_frequency, by = c("iso_a3" = "iso3")) %>%
    filter(continent != "Antarctica") %>%
    tm_shape() +
    tm_polygons(
        fill = "n",
        fill.scale = tm_scale_intervals(
            breaks = c(1, 100, 250, 500, Inf),
            values.range = c(0.1, 0.9),
            value.na = "grey80",
            label.na = "No Data"
        ),
        fill.legend = tm_legend(title = "No. of data points"),
        lwd = 0.5
    ) +
    tm_title("JMP raw data collection - Total number of data points since 2000") +
    tm_layout(scale = 1.5)

top10(jmp_iso3_frequency)
```

## Data points for user interface since 2015

```{r}

world_moll %>%
    left_join(jmp_iso3_user_interface_2015, by = c("iso_a3" = "iso3")) %>%
    filter(continent != "Antarctica") %>%
    tm_shape() +
    tm_polygons(
        fill = "n",
        fill.scale = tm_scale_intervals(
            breaks = c(1, 11, 21, 31, Inf),
            values.range = c(0.1, 0.9),
            value.na = "grey80",
            label.na = "No Data"
        ),
        fill.legend = tm_legend(title = "No. of data points"),
        lwd = 0.5
    ) +
    tm_title("JMP raw data collection - Number of data points for 'user interface' since 2015") +
    tm_layout(scale = 1.5)

top10(jmp_iso3_user_interface_2015)

```

## Data points for emptying since 2015

```{r}

world_moll %>%
    left_join(jmp_iso3_emptying_2015, by = c("iso_a3" = "iso3")) %>%
    filter(continent != "Antarctica") %>%
    tm_shape() +
    tm_polygons(
        fill = "n",
        fill.scale = tm_scale_intervals(
            breaks = c(1, 11, 21, 31, Inf),
            values.range = c(0.1, 1),
            value.na = "grey90",
            label.na = "No Data"
        ),
        fill.legend = tm_legend(title = "No. of data points"),
        lwd = 0.5
    ) +
    tm_title("JMP raw data collection - Number of data points for 'emptying' since 2015") +
    tm_layout(scale = 1.5)

top10(jmp_iso3_emptying_2015)

```

## Data points for wastewater treatment since 2015

```{r}

world_moll %>%
    left_join(jmp_iso3_ww_treatment_2015, by = c("iso_a3" = "iso3")) %>%
    filter(continent != "Antarctica") %>%
    tm_shape() +
    tm_polygons(
        fill = "n",
        fill.scale = tm_scale_intervals(
            breaks = c(1, 3, 6, 20, Inf),
            values.range = c(0.1, 1),
            value.na = "grey90",
            label.na = "No Data"
        ),
        fill.legend = tm_legend(title = "No. of data points"),
        lwd = 0.5
    ) +
    tm_title("JMP raw data collection - Number of data points for 'wastewater treatment' since 2015") +
    tm_layout(scale = 1.5)

top10(jmp_iso3_ww_treatment_2015)

```

## Data points for faecal sludge treatment since 2015

```{r}

world_moll %>%
    left_join(jmp_iso3_fs_treatment_2015, by = c("iso_a3" = "iso3")) %>%
    filter(continent != "Antarctica") %>%
    tm_shape() +
    tm_polygons(
        fill = "n",
        fill.scale = tm_scale_intervals(
            breaks = c(1, 3, 6, 20, Inf),
            values.range = c(0.1, 1),
            value.na = "grey90",
            label.na = "No Data"
        ),
        fill.legend = tm_legend(title = "No. of data points"),
        lwd = 0.5
    ) +
    tm_title("JMP raw data collection - Number of data points for 'faecal sludge' since 2015") +
    tm_layout(scale = 1.5)

top10(jmp_iso3_fs_treatment_2015)


```
